Optimal updating magnitude in adaptive flat-distribution-sampling simulations

### 2017-06-23 doc/invt.tex
small changes

Free energy calculation\cite{frenkel, newman} is a central theme
in computational physics and chemistry
that can provide insight into an array of phenomena not easily studied
with traditional experiments.
%
Given a system,
often the task is to compute
a distribution, $p(z)$,
along a collective variable, $z$, by sampling via either Monte Carlo\cite{
  frenkel, newman, landau_binder} (MC)
or molecular dynamics\cite{frenkel, karplus2002} (MD) simulations.
%
The dimensionless free energy, or potential of mean force (PMF),
along the collective variable, $z$,
is $-\ln p(z)$.
%
However, accurately estimating the PMF is often difficult
when the collective variable is energetically restricted to local regions
of the complex, multimodal free energy surface.
%
Thus,
to overcome these energetic barriers and
to capture the global shape of the free energy surface
an effective strategy is to introduce a bias potential that
cancels the PMF.
%
Ideally, then, sampling yields a flat distribution
along $z$\cite{mezei1987, berg1992, *lee1993,
wang2001, *wang2001pre,
huber1994,
*laio2002, *laio2008, *barducci2011, *sutto2012}.
%
The flat-distribution sampling can be used to accelerate
sampling of slow transitions that otherwise would not readily
be observed with traditional simulations.
%
%A better solution is to artificially flatten
%the target distribution along $z$\cite{mezei1987, berg1992, *lee1993,
%wang2001, *wang2001pre,
%huber1994,
%*laio2002, *laio2008, *barducci2011, *sutto2012}.
%%
%This alleviates the above problem of local trapping
%by forcing the system to travel more
%deliberately along $z$,
%which often represents
%a direction of slow motion.
%%
%To achieve a flat distribution, however,
%the method must be able to construct a
%bias potential that cancels the PMF.

Many efficient flat-distribution-sampling (FDS) techniques
based on adaptive construction of the bias potential
have been introduced,
including the Wang-Landau (WL) algorithm\cite{
  wang2001, *wang2001pre}
and metadynamics\cite{huber1994,
  *laio2002, *laio2008, *barducci2011, *sutto2012, micheletti2004}
among others\cite{kim2006, *kim2007, junghans2014,
  langfeld2012, pellegrini2014}.
%
%flat-distribution-sampling (FDS) techniques
%that build the bias potential adaptively or on the fly.
%
These techniques regularly update the bias potential
to discourage future visits to previously sampled configurations
by incrementally elevating the bias potential,
%and they are closely-related\cite{junghans2014, micheletti2004}.
%
%The bias potential is associated with discrete values of $z$.
%
A key difference lies
in the updating window function,
which specifies
a neighborhood around the current value of
$z$ on the free energy surface
where the updating should occur
as well as the relative updating magnitude.
%
Usually, the window function
takes the form of a discrete
$\delta$-function (a boxcar function one histogram bin wide)
in WL,
or that of a Gaussian
in metadynamics.
%
We will refer to the two updating schemes
as the single-bin
and Gaussian updating schemes, respectively.
%
While the former is popular for MC simulations\cite{wang2001,
  wang2001pre, kim2006, *kim2007},
the latter is more popular for MD simulations
as it provides a smoother bias potential
that can be readily differentiated for force calculation.\cite{huber1994,
  *laio2002, *laio2008, *barducci2011, *sutto2012, junghans2014}
%
%is that in the WL case,
%each updating step is limited to the bin
%containing the current $z$,
%while metadynamics adopts an extended
%Gaussian window
%covering several neighboring bins.
%%
%The latter is more often used in MD simulations
%as it guarantees a smooth profile
%of the bias potential
%suitable for numerical differentiation
%in deriving the bias force.
%
In both cases,
regular updates to the bias potential
%in both cases, the bias potential
%is updated frequently (usually every few MC or MD steps),
%and the fluctuation
disrupts the underlying equilibrium
sampling\cite{zhou2005, morozov2007, zhou2008},
and for convergence of the estimated PMF
one  has to gradually decrease
the updating magnitude of the bias.

@@

Many methods\cite{mezei1987, berg1992, *lee1993,
wang2001, *wang2001pre, huber1994,
*laio2002, *laio2008, *barducci2011, *sutto2012}
have been developed to find the proper bias potential.
%
In a class of equilibrium FDS methods\cite{
  mezei1987, berg1992, *lee1993, marinari1992, *lyubartsev1992},
the bias potential, $u_i$, is
estimated a priori and fixed
during simulation,
%
and a correction to the bias potential
is derived from the normalized histogram, $H_i$,
accumulated from the simulation as
%
%Using $H_i$ as an estimate of $\pi_i$ in
%Eq. \eqref{eq:pi_p_V}
%allows us to correct $u_i$ as
%
\begin{equation}
  \hat u_i
  =
  u_i
  +
  \ln \frac{ H_i }
           { \rho_i }.
  \label{eq:vcorr_equil}
\end{equation}
%
\note{This follows from
  \begin{align*}
    H_i \approx \pi_i
    &\propto p_i \, e^{-u_i},
    \\
    \rho_i
    &\propto p_i \, e^{-\hat u_i}.
  \end{align*}
}
However, this formula often requires a long
accumulation period for a precise histogram,
and thus disallows
continuous improvement of the bias potential
%(hence that of the PMF)
as simulation lengthens.

An other class of adaptive FDS methods,
including the WL algorithm\cite{wang2001, *wang2001pre}
and metadynamics\cite{huber1994, laio2002, *laio2008, *barducci2011, *sutto2012},
address the above shortcoming
by continuously and incrementally updating the bias potential,
and we will focus on these methods below.
%%
%%These methods incrementally update the bias potential
%%and can approximately sample the target distribution.
%%
%However,
%the on-the-fly updates break the microscopic reversibility
%of the underlying sampling mechanism.
%Further, failure to properly reduce the magnitude of
%the histogram-based increments
%would lead to roughness or errors in the final estimate of the PMF.
%%
%%Thus, for a conservative practitioner,
%%the E-FDS methods appear more appealing
%%once a sufficiently accurate bias potential is available.
%%
%To minimize this effect,
%many adaptive methods adopt a schedule to
%reduce the updating magnitude
%over the course of simulation\cite{
%marsili2006,
%belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
%liang2007,
%barducci2008}.
%%
%%Thus, in late stages of a long simulation,
%%the updating becomes negligible and
%%the adaptive simulation tends to
%%the corresponding equilibrium one.
%%
%%if the bias potential were converged to the exact value,
%%this adaptive FDS tends to the ideal equilibrium counterpart,
%%particular equilibrium simulation can be thought as
%%the optimal asymptotic limit of the adaptive counterpart,
%%
%For example, one can show that
%if the magnitude is controlled by
%the optimal inverse-time prescription\cite{
%belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
%liang2007,
%morozov2007, zhou2008,
%komura2012, *caparica2012, *caparica2014},
%the WL algorithm is equivalent to the ideal equilibrium counterpart
%in improving the precision of the bias potential,
%thus allowing for maximum efficiency
%(cf. Appendix \ref{sec:equilerr}).
%%the bias potential in the WL algorithm
%%will evolve similar to Eq. \eqref{eq:vcorr_equil}
%%with the precision of the histogram as simulation lengthens.
%
%
%\subsubsection{Updating schemes}
%
%
%
%We will mainly focus on the adaptive methods below.
%
In the WL algorithm\cite{wang2001, *wang2001pre},
the bias potential, $u_i(t)$, is updated
at each sampling step $t$ as
%
\begin{equation}
  u_i(t+1)
  =
  u_i(t)
  +
  \delta_{i, \, i(t)}
  \frac{ \alpha(t) } { \rho_{i(t)} }
  ,
\label{eq:wl_update}
\end{equation}
%
where $i(t)$ is the bin at step $t$,
$\delta_{i, \, i(t)}$ is the Kronecker delta function,
which is $1$ when $i = i(t)$ or $0$ otherwise,
and $\alpha(t)$ is the updating magnitude.
%
We refer to this scheme as the \emph{single-bin scheme},
for it applies only to the bias potential
at the current bin, $i(t)$.
%
In a more general \emph{multiple-bin scheme},
%including the Gaussian updating scheme used in metadynamics,
the updating also involves neighboring bins:
%
\begin{equation}
  u_i(t+1)
  =
  u_i(t)
  +
  w_{i, \, i(t)}
  \frac{ \alpha(t) }
       { \rho_{i(t)} }
  .
  \label{eq:mbin_update}
\end{equation}
%
The Gaussian updating scheme used in metadynamics
is an example, with $w_{ij}$ being a Gaussian of $i-j$.
%
Note that if the updating occurs
only every few sampling steps,
then $t$ means the number of updating steps.

@@

In the original WL papers\cite{
wang2001, *wang2001pre},
the updating magnitude, $\alpha(t)$,
(or $\ln f$ therein)
is controlled in stages.
%
In each stage, $\alpha(t)$
is kept constant,
and the histogram along $i$
is collected and monitored.
%
Once the histogram is sufficiently flat,
we switch to a new stage
with $\alpha(t)$ reduced by a factor of $1/2$\cite{
wang2001, *wang2001pre}.
%
While this scheme works well for early stages,
it tends to reduce the magnitude
too quickly in late stages, making the asymptotic error
saturate\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016}.

@@

We can now use Eq. \eqref{eq:eig_w} to diagonalize
the multiple-bin updating scheme, Eq. \eqref{eq:mbin_update},
as
%
$$
{\tilde v}_k(t + 1) =
{\tilde v}_k(t) + \alpha(t) \, \lambda_k \,
{\tilde f}_k(t)
,
$$
or in continuous time, we have, for $k > 0$,
%
\begin{equation}
  \dot{\tilde v}_k(t)
  =
  \alpha(t) \, \lambda_k \, {\tilde f}_k(t)
  \approx
  -\alpha(t) \, \lambda_k \,
  \bigl[ {\tilde v}_k(t) - {\tilde \zeta}_k(t) \bigr]
  ,
  \label{eq:vt_diffeq_mbin}
\end{equation}
%
where in the second step,
we have used Eq. \eqref{eq:xtstar} and
the transformed version of Eq. \eqref{eq:sh_ave}
by assuming the linear approximation
as in Sec. \ref{sec:sbin_diffeq}.

@@

We now determine the optimal value of $q(T)$
that minimizes the total error defined in Eq. \eqref{eq:error_tot}.
%
We will consider simulations in which
the system is initially equilibrated at a
constant updating magnitude, $a_0$.
%
From Eqs. \eqref{eq:error_res}, \eqref{eq:xt2_eql1},
and \eqref{eq:error_asym2},
we find that
the optimal $q(T)$ satisfies
%
\begin{align}
  \mathcal Y\bigl( q(T) \bigr)
  &\equiv
  \frac{ M\bigl( q(T) \bigr) } { T }
    \int_0^{q(T)} M(q') \, d q'
  -\frac{a_0}{2} \, M^2\bigl( q(T) \bigr)
  \notag \\
  &
  -\sum_{k = 1}^{n-1}
  \lambda_k \, \epsilon_k \, e^{-2\lambda_k \, q(T)}
  =0
  .
\label{eq:opt_qT}
\end{align}
%
%\begin{align}
%  m\bigl( q(T) \bigr)
%  +
%  \frac{2}{a_0 \, C_M \, M\bigl( q(T) \bigr)}
%  \sum_{k = 1}^{n-1}
%  \lambda_k \, \tilde u_k^2 \,
%  e^{-2 \, \lambda_k \, [a_0 \, T_0 + q(T)] }
%  \\=
%  \frac{1} { T \, (a_0 / 2) }
%  .
%\label{eq:opt_qT}
%\end{align}
%
\note{Let $q_T \equiv q(T)$,
$$
\begin{aligned}
  \frac{
    \partial \Err_R(T)
  }
  {
    \partial q_T
  }
  &=
  -a_0 \, M^2(q_T)
  -2
  \sum_{k=1}^{n-1} \lambda_k \,
  \tilde u_k^2 e^{-2 \, \lambda_k \, (a_0 \, T_0 + q_T)}
  ,
  \\
  \frac{
    \partial \Err_A(T)
  }
  {
    \partial q_T
  }
  &=
  \frac 2 T \,
  M(q_T) \,
  \int_0^{ q_T } M(q) \, dq
  =
  \frac{ 2 \, C_M } { T } \, M(q_T)
  .
\end{aligned}
$$
Then solve $\partial (\Err_R + \Err_A) / \partial Q_c = 0$.
}
%
We can rewrite the condition in terms of
the initial updating magnitude by Eq. \eqref{eq:mQ_invTa},
\begin{align}
  \alpha(0)
  =
  \frac{1}{T \, m\bigl( q(T) \bigr) }
  =
  \frac{ a_0 } { 2 }
  +
  \sum_{k = 1}^{n-1}
  \frac{
    \lambda_k \, \epsilon_k \, e^{-2\lambda_k \, q(T)}
  }{M^2\bigl( q(T) \bigr)}
  .
  \label{eq:opt_alpha0}
\end{align}
%
If the initial bias, $\epsilon_k$, is negligible,
which is typically true if no eigenvalue is close to $0$,
then
%
\begin{equation}
  \alpha( t = 0 )
  \approx
  \frac{ a_0 }
       { 2 }
  ,
\label{eq:half_alpha0}
\end{equation}
%
i.e. the optimal initial updating magnitude
is half of the equilibrium value.\footnote{This factor, $1/2$,
happens to coincide with the
recommended reduction factor of the updating magnitude
at stage transitions
in the original WL algorithm\cite{
wang2001, *wang2001pre}.
}

@@

\item \label{step:alpha}
Compute the optimal schedule by
  numerically integrating Eqs. \eqref{eq:q_opt} and \eqref{eq:mint}.
%
To handle Gaussian updating schemes more effectively,
  we have chosen the $\mathcal N+1$ grid points over $[0, q(T)]$ to be given by
  $q_{[i]} = \bigl(q(T) + 1\bigr) - \bigl(q(T)+1\bigr)^{1-i/\mathcal N}$,
  where $i = 0, \dots, \mathcal N$, and we chose $\mathcal N=20000$.
%
The curve $q(t)$ is then integrated
  using the trapezoidal rule\cite{press3rd},
  resulting in a list of values, $\bigl(q_{[i]}, t_{[i]} \bigr)$.
%
We then compute the updating magnitude by numerical differentiation:
  $\alpha_i = (1 - e^{-\lambda_1 \, \hat \alpha_i})/\lambda_1$
  and
  $\hat \alpha_i \equiv (q_{[i+1]} - q_{[i-1]}) / (t_{[i+1]} - t_{[i-1]})$,
  with
  $q_{[ -1]} \equiv 0$,
  $q_{[N+1]} \equiv q(T)$,
  $t_{[ -1]} \equiv 0$, and
  $t_{[N+1]} \equiv T$.
%
The resulting values of $(t_{[i]}, \alpha_{[i]})$
  can be linear interpolated
  to get the updating magnitude, $\alpha(t)$, at any simulation time, $t$.

@@

The above procedure is not always necessary.
%
As shown below, for a class of bandpass updating schemes,
  which resemble the Gaussian updating schemes,
  the asymptotically optimal schedule
  is simply given by the inverse-time schedule, Eq. \eqref{eq:alpha_invt1}.
%
For these updating schemes,
  the preliminary run with a constant updating magntiude
  can be replaced by the WL-style stagewise reduction
  until the updating magnitude falls under
  the value given the inverse-time prescription\cite{
    belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016}.

@@

In addition to the application to homogeneous updating schemes
(Appendix \ref{sec:homo_bandpass}),
bandpass updating schemes can also be applied to
a continuous variable, $z$,\footnote{
  In the transition to a continuous variable,
  we will map
  $\rho_i$ to $\rho(z) \delta z$,
  $\phi_{ki}$ to $\phi_k(z) \delta z$,
  $w_{ij}$ to $w(z, z') \delta z$, etc.}
with the eigenmodes derived from orthogonal polynomials, $R_k(z)$,
that satisfy
\begin{align*}
  \int R_k(z) \, R_{k'}(z) \, \rho(z) \, d z = \delta_{k, k'}
  .
\end{align*}
%
With $\phi_k(z) = R_k(z) \, \rho(z)$,
we can readily satisfy Eqs. \eqref{eq:eig_orthonormal_cols}-\eqref{eq:ortho0},
and construct the updating matrix by
Eqs. \eqref{eq:w_from_phi} and \eqref{eq:lambda_bandpass} as
\begin{align*}
  w(z, z')
  =
  \sum_{k=1}^{K-1} \frac{ \phi_k(z) \, \phi_k(z') } { \rho(z) }
  = \sum_{k=1}^{K-1} R_k(z) \, R_k(z') \, \rho(z'),
\end{align*}
%
where we have dropped the $k=0$ mode
that corresponds to the uniform shift of the bias potential.
%
This results in a polynomial bias potential
that can be written as a superposition
$$
u(z, t) = \sum_{k=1}^{K-1} c_k(t) \, R_k(z),
$$
with the coefficients updated as, according to Eq. \eqref{eq:mbin_update},
$$
c_k(t+1) = c_k(t) + \alpha(t) \, R_k\bigl( z(t) \bigr).
$$

For example,
to sample a flat distribution, $\rho(z)$,
over $[z_c - \Delta z/2, z_c + \Delta z/2]$,
we can use the Legendre polynomials\cite{arfken},
and with $K = 2$,
the bias potential contains only a linear mode for
$R_1(z) = 2 \sqrt{3} \, (z - z_c)/\Delta z$,
which recovers the Langfeld, Lucini and Rago (LLR)
algorithm\cite{langfeld2012}.
%
Applied to the energy space, $z = E$,
$c_1(t)/\sigma_z$ represents the scaled local or
statistical temperature\cite{kim2006, *kim2007},
and the inverse-time schedule
was recommended for its update\cite{pellegrini2014}.

>>>>

Free energy calculation\cite{frenkel, newman} is a central theme
in computational physics and chemistry
that can provide insight into an array of phenomena not easily studied
with traditional experiments.
%
Given a system,
often the task is to compute
a distribution, $p(z)$,
along a collective variable, $z$, by sampling via either Monte Carlo\cite{
  frenkel, newman, landau_binder} (MC)
or molecular dynamics\cite{frenkel, karplus2002} (MD) simulations.
%
The dimensionless free energy, or potential of mean force (PMF),
along the collective variable, $z$,
is $-\ln p(z)$.
%
However, accurately estimating the PMF is often difficult
when the collective variable is energetically restricted to local regions
of the complex, multimodal free energy surface.
%
Thus,
to overcome these energetic barriers and
to capture the global shape of the free energy surface
an effective strategy is to introduce a bias potential that
cancels the PMF.
%
Ideally, then, sampling yields a flat distribution
along $z$\cite{mezei1987, berg1992, *lee1993,
wang2001, wang2001pre,
huber1994,
*laio2002, *laio2008, *barducci2011, *sutto2012}.
%
The flat-distribution sampling can be used to accelerate
sampling of slow transitions that otherwise would not readily
be observed with traditional simulations.
%
%A better solution is to artificially flatten
%the target distribution along $z$\cite{mezei1987, berg1992, *lee1993,
%wang2001, wang2001pre,
%huber1994,
%*laio2002, *laio2008, *barducci2011, *sutto2012}.
%%
%This alleviates the above problem of local trapping
%by forcing the system to travel more
%deliberately along $z$,
%which often represents
%a direction of slow motion.
%%
%To achieve a flat distribution, however,
%the method must be able to construct a
%bias potential that cancels the PMF.

Many efficient flat-distribution-sampling (FDS) techniques
based on adaptive construction of the bias potential
have been introduced,
including the Wang-Landau (WL) algorithm\cite{
  wang2001, wang2001pre}
and metadynamics\cite{huber1994,
  *laio2002, *laio2008, *barducci2011, *sutto2012, micheletti2004}
among others\cite{kim2006, *kim2007, junghans2014,
  langfeld2012, pellegrini2014}.
%
%flat-distribution-sampling (FDS) techniques
%that build the bias potential adaptively or on the fly.
%
These techniques regularly update the bias potential
to discourage future visits to previously sampled configurations
by incrementally elevating the bias potential,
%and they are closely-related\cite{junghans2014, micheletti2004}.
%
%The bias potential is associated with discrete values of $z$.
%
A key difference lies
in the updating window function,
which specifies
a neighborhood around the current value of
$z$ on the free energy surface
where the updating should occur
as well as the relative updating magnitude.
%
Usually, the window function
takes the form of a discrete
$\delta$-function (a boxcar function one histogram bin wide)
in WL,
or that of a Gaussian
in metadynamics.
%
We will refer to the two updating schemes
as the single-bin
and Gaussian updating schemes, respectively.
%
While the former is popular for MC simulations\cite{wang2001,
  wang2001pre, kim2006, *kim2007},
the latter is more popular for MD simulations
as it provides a smoother bias potential
that can be readily differentiated for force calculation.\cite{huber1994,
  *laio2002, *laio2008, *barducci2011, *sutto2012, junghans2014}
%
%is that in the WL case,
%each updating step is limited to the bin
%containing the current $z$,
%while metadynamics adopts an extended
%Gaussian window
%covering several neighboring bins.
%%
%The latter is more often used in MD simulations
%as it guarantees a smooth profile
%of the bias potential
%suitable for numerical differentiation
%in deriving the bias force.
%
In both cases,
regular updates to the bias potential
%in both cases, the bias potential
%is updated frequently (usually every few MC or MD steps),
%and the fluctuation
disrupts the underlying equilibrium
sampling\cite{zhou2005, morozov2007, zhou2008},
and for convergence of the estimated PMF
one  has to gradually decrease
the updating magnitude of the bias.

@@

Many methods\cite{mezei1987, berg1992, *lee1993,
wang2001, wang2001pre, huber1994,
*laio2002, *laio2008, *barducci2011, *sutto2012}
have been developed to find the proper bias potential.
%
In a class of equilibrium FDS methods\cite{
  mezei1987, berg1992, *lee1993, marinari1992, *lyubartsev1992},
the bias potential, $u_i$, is
estimated a priori and fixed
during simulation,
%
and a correction to the bias potential
is derived from the normalized histogram, $H_i$,
accumulated from the simulation as
%
%Using $H_i$ as an estimate of $\pi_i$ in
%Eq. \eqref{eq:pi_p_V}
%allows us to correct $u_i$ as
%
\begin{equation}
  \hat u_i
  =
  u_i
  +
  \ln \frac{ H_i }
           { \rho_i }.
  \label{eq:vcorr_equil}
\end{equation}
%
\note{This follows from
  \begin{align*}
    H_i \approx \pi_i
    &\propto p_i \, e^{-u_i},
    \\
    \rho_i
    &\propto p_i \, e^{-\hat u_i}.
  \end{align*}
}
However, this formula often requires a long
accumulation period for a precise histogram,
and thus disallows
continuous improvement of the bias potential
%(hence that of the PMF)
as simulation lengthens.

An other class of adaptive FDS methods,
including the WL algorithm\cite{wang2001, wang2001pre}
and metadynamics\cite{huber1994, laio2002, *laio2008, *barducci2011, *sutto2012},
address the above shortcoming
by continuously and incrementally updating the bias potential,
and we will focus on these methods below.
%%
%%These methods incrementally update the bias potential
%%and can approximately sample the target distribution.
%%
%However,
%the on-the-fly updates break the microscopic reversibility
%of the underlying sampling mechanism.
%Further, failure to properly reduce the magnitude of
%the histogram-based increments
%would lead to roughness or errors in the final estimate of the PMF.
%%
%%Thus, for a conservative practitioner,
%%the E-FDS methods appear more appealing
%%once a sufficiently accurate bias potential is available.
%%
%To minimize this effect,
%many adaptive methods adopt a schedule to
%reduce the updating magnitude
%over the course of simulation\cite{
%marsili2006,
%belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
%liang2007,
%barducci2008}.
%%
%%Thus, in late stages of a long simulation,
%%the updating becomes negligible and
%%the adaptive simulation tends to
%%the corresponding equilibrium one.
%%
%%if the bias potential were converged to the exact value,
%%this adaptive FDS tends to the ideal equilibrium counterpart,
%%particular equilibrium simulation can be thought as
%%the optimal asymptotic limit of the adaptive counterpart,
%%
%For example, one can show that
%if the magnitude is controlled by
%the optimal inverse-time prescription\cite{
%belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
%liang2007,
%morozov2007, zhou2008,
%komura2012, *caparica2012, *caparica2014},
%the WL algorithm is equivalent to the ideal equilibrium counterpart
%in improving the precision of the bias potential,
%thus allowing for maximum efficiency
%(cf. Appendix \ref{sec:equilerr}).
%%the bias potential in the WL algorithm
%%will evolve similar to Eq. \eqref{eq:vcorr_equil}
%%with the precision of the histogram as simulation lengthens.
%
%
%\subsubsection{Updating schemes}
%
%
%
%We will mainly focus on the adaptive methods below.
%
In the WL algorithm\cite{wang2001, wang2001pre},
the bias potential, $u_i(t)$, is updated
at each sampling step $t$ as
%
\begin{equation}
  u_i(t+1)
  =
  u_i(t)
  +
  \delta_{i, \, i(t)}
  \frac{ \alpha(t) } { \rho_{i(t)} }
  ,
\label{eq:wl_update}
\end{equation}
%
where $i(t)$ is the bin at step $t$,
$\delta_{i, \, i(t)}$ is the Kronecker delta function,
which is $1$ when $i = i(t)$ or $0$ otherwise,
and $\alpha(t)$ is the updating magnitude.
%
We refer to this scheme as the \emph{single-bin scheme},
for it applies only to the bias potential
at the current bin, $i(t)$.
%
In a more general \emph{multiple-bin scheme},
%including the Gaussian updating scheme used in metadynamics,
the updating also involves neighboring bins:
%
\begin{equation}
  u_i(t+1)
  =
  u_i(t)
  +
  w_{i, \, i(t)}
  \frac{ \alpha(t) }
       { \rho_{i(t)} }
  .
  \label{eq:mbin_update}
\end{equation}
%
The Gaussian updating scheme used in metadynamics
is an example, with $w_{ij}$ being a Gaussian of $i-j$.
%
Note that if the updating occurs
only every few sampling steps,
then $t$ means the number of updating steps.

@@

In the original WL algorithm\cite{
wang2001, wang2001pre},
the updating magnitude, $\alpha(t)$,
(or $\ln f$ therein)
is controlled in stages.
%
In each stage, $\alpha(t)$
is kept constant,
and the histogram along $i$
is collected and monitored.
%
Once the histogram is sufficiently flat,
we switch to a new stage
with $\alpha(t)$ reduced by a factor of $1/2$\cite{
wang2001, wang2001pre}.
%
While this scheme works well for early stages,
it tends to reduce the magnitude
too quickly in late stages, making the asymptotic error
saturate\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016}.

@@

We can now use Eq. \eqref{eq:eig_w} to diagonalize
the multiple-bin updating scheme, Eq. \eqref{eq:mbin_update},
as
%
\begin{equation}
  {\tilde v}_k(t + 1) =
  {\tilde v}_k(t) + \alpha(t) \, \lambda_k \,
  {\tilde f}_k(t)
  ,
  \label{eq:vkupdate}
\end{equation}
%
or in continuous time, we have, for $k > 0$,
%
\begin{equation}
  \dot{\tilde v}_k(t)
  =
  \alpha(t) \, \lambda_k \, {\tilde f}_k(t)
  \approx
  -\alpha(t) \, \lambda_k \,
  \bigl[ {\tilde v}_k(t) - {\tilde \zeta}_k(t) \bigr]
  ,
  \label{eq:vt_diffeq_mbin}
\end{equation}
%
where in the second step,
we have used Eq. \eqref{eq:xtstar} and
the transformed version of Eq. \eqref{eq:sh_ave}
by assuming the linear approximation
as in Sec. \ref{sec:sbin_diffeq}.

@@

We now determine the optimal value of $q(T)$
that minimizes the total error defined in Eq. \eqref{eq:error_tot}.
%
We will consider simulations in which
the system is initially equilibrated at a
constant updating magnitude, $a_0$.
%
From Eqs. \eqref{eq:error_res}, \eqref{eq:xt2_eql1},
and \eqref{eq:error_asym2},
we find that
the optimal $q(T)$ satisfies
%
\begin{align}
  \mathcal Y\bigl( q(T) \bigr)
  &\equiv
  \frac{ M\bigl( q(T) \bigr) } { T }
    \int_0^{q(T)} M(q') \, d q'
  -\frac{a_0}{2} \, M^2\bigl( q(T) \bigr)
  \notag \\
  &
  -\sum_{k = 1}^{n-1}
  \lambda_k \, \epsilon_k \, e^{-2\lambda_k \, q(T)}
  =0
  .
\label{eq:opt_qT}
\end{align}
%
%\begin{align}
%  m\bigl( q(T) \bigr)
%  +
%  \frac{2}{a_0 \, C_M \, M\bigl( q(T) \bigr)}
%  \sum_{k = 1}^{n-1}
%  \lambda_k \, \tilde u_k^2 \,
%  e^{-2 \, \lambda_k \, [a_0 \, T_0 + q(T)] }
%  \\=
%  \frac{1} { T \, (a_0 / 2) }
%  .
%\label{eq:opt_qT}
%\end{align}
%
\note{Let $q_T \equiv q(T)$,
$$
\begin{aligned}
  \frac{
    \partial \Err_R(T)
  }
  {
    \partial q_T
  }
  &=
  -a_0 \, M^2(q_T)
  -2
  \sum_{k=1}^{n-1} \lambda_k \,
  \tilde u_k^2 e^{-2 \, \lambda_k \, (a_0 \, T_0 + q_T)}
  ,
  \\
  \frac{
    \partial \Err_A(T)
  }
  {
    \partial q_T
  }
  &=
  \frac 2 T \,
  M(q_T) \,
  \int_0^{ q_T } M(q) \, dq
  =
  \frac{ 2 \, C_M } { T } \, M(q_T)
  .
\end{aligned}
$$
Then solve $\partial (\Err_R + \Err_A) / \partial Q_c = 0$.
}
%
We can rewrite the condition in terms of
the initial updating magnitude by Eq. \eqref{eq:mQ_invTa},
\begin{align}
  \alpha(0)
  =
  \frac{1}{T \, m\bigl( q(T) \bigr) }
  =
  \frac{ a_0 } { 2 }
  +
  \sum_{k = 1}^{n-1}
  \frac{
    \lambda_k \, \epsilon_k \, e^{-2\lambda_k \, q(T)}
  }{M^2\bigl( q(T) \bigr)}
  .
  \label{eq:opt_alpha0}
\end{align}
%
If the initial bias, $\epsilon_k$, is negligible,
which is typically true if no eigenvalue is close to $0$,
then
%
\begin{equation}
  \alpha( t = 0 )
  \approx
  \frac{ a_0 }
       { 2 }
  ,
\label{eq:half_alpha0}
\end{equation}
%
i.e. the optimal initial updating magnitude
is half of the equilibrium value.\footnote{This factor, $1/2$,
happens to coincide with the
recommended reduction factor of the updating magnitude
at stage transitions
in the original WL algorithm\cite{
wang2001, wang2001pre}.
}

@@

\item \label{step:alpha}
Compute the optimal schedule by
  numerically integrating Eqs. \eqref{eq:q_opt} and \eqref{eq:mint}.
%
To handle Gaussian updating schemes more effectively,
  we have chosen the $\mathcal N+1$ grid points over $[0, q(T)]$ to be given by
  $q_{[i]} = \bigl(q(T) + 1\bigr) - \bigl(q(T)+1\bigr)^{1-i/\mathcal N}$,
  where $i = 0, \dots, \mathcal N$, and we chose $\mathcal N=20000$.
%
The curve $q(t)$ is then integrated
  using the trapezoidal rule\cite{press3rd},
  resulting in a list of values, $\bigl(q_{[i]}, t_{[i]} \bigr)$.
%
We then compute the updating magnitude by numerical differentiation:
  $\alpha_i = (1 - e^{-\lambda_1 \, \hat \alpha_i})/\lambda_1$\footnote{This
    discretization follows from runtime average interpretation
    discussed in Appendix \ref{sec:equilerr}.}
  and
  $\hat \alpha_i \equiv (q_{[i+1]} - q_{[i-1]}) / (t_{[i+1]} - t_{[i-1]})$,
  with
  $q_{[ -1]} \equiv 0$,
  $q_{[N+1]} \equiv q(T)$,
  $t_{[ -1]} \equiv 0$, and
  $t_{[N+1]} \equiv T$.
%
The resulting values of $(t_{[i]}, \alpha_{[i]})$
  can be linear interpolated
  to get the updating magnitude, $\alpha(t)$, at any simulation time, $t$.

@@

The above procedure is not always necessary.
%
As shown below, for a class of bandpass updating schemes,
  which resemble the Gaussian updating schemes,
  the asymptotically optimal schedule
  is simply given by the inverse-time schedule, Eq. \eqref{eq:alpha_invt1}.
%
%For these updating schemes,
%  the preliminary run with a constant updating magntiude
%  can be replaced by the WL-style stagewise reduction
%  until the updating magnitude falls under
%  the value given the inverse-time prescription\cite{
%    belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016}.

@@

In addition to the application to homogeneous updating schemes
(Appendix \ref{sec:homo_bandpass}),
bandpass updating schemes can also be applied to
a continuous variable, $z$,\footnote{
  In the transition to a continuous variable,
  we will map
  $\rho_i$ to $\rho(z) \delta z$,
  $\phi_{ki}$ to $\phi_k(z) \delta z$,
  $w_{ij}$ to $w(z, z') \delta z$, etc.}
with the eigenmodes derived from orthogonal polynomials, $R_k(z)$,
that satisfy
\begin{align*}
  \int R_k(z) \, R_{k'}(z) \, \rho(z) \, d z = \delta_{k, k'}
  .
\end{align*}
%
With $\phi_k(z) = R_k(z) \, \rho(z)$,
we can readily satisfy Eqs. \eqref{eq:eig_orthonormal_cols}-\eqref{eq:ortho0},
and construct the updating matrix by
Eqs. \eqref{eq:w_from_phi} and \eqref{eq:lambda_bandpass} as
\begin{align*}
  w(z, z')
  =
  \sum_{k=1}^{K-1} \frac{ \phi_k(z) \, \phi_k(z') } { \rho(z) }
  = \sum_{k=1}^{K-1} R_k(z) \, R_k(z') \, \rho(z'),
\end{align*}
%
where we have dropped the $k=0$ mode
that corresponds to a uniform shift
of the bias potential.
%
The resulting bias potential
assumes a polynomial form
and can be written as a superposition
$u(z, t) = \sum_{k=1}^{K-1} c_k(t) \, R_k(z)$,
with the coefficients updated as, according to Eq. \eqref{eq:mbin_update},
\begin{equation}
  c_k(t+1) = c_k(t) + \alpha(t) \, R_k\bigl( z(t) \bigr)
  .
  \label{eq:ckupdate}
\end{equation}

For example,
to sample a flat distribution, $\rho(z)$,
over $[z_c - \Delta z/2, z_c + \Delta z/2]$,
we can use the Legendre polynomials\cite{arfken},
and with $K = 2$,
the bias potential contains only a linear mode for
$R_1(z) = 2 \sqrt{3} \, (z - z_c)/\Delta z$,
with its coefficient, $c_1$, updated according to
the difference between the average of $z$
and the center $z_c$.
%
We thus recover the Langfeld-Lucini-Rago (LLR)
algorithm\cite{langfeld2012},
and the inverse-time schedule
was recommended for optimal convergence of $c_1$\cite{pellegrini2014}.
